---
title: "AK - Course Project -1"
output: html_document
---

##Setting
Let's make sure settings are correct we bore we proceed with the project.
```{r setup, }
knitr::opts_chunk$set(echo = TRUE)
```

##Loading and preprocessing the data
Let's load data file into R and see how first ten rows look like.

```{r load,}
mydata<-read.csv("activity.csv",header = TRUE,sep = ",")
head(mydata,10)
```


##What is mean total number of steps taken per day?

###Ignoring missing values, calculate the total number of steps taken per day.
Let's see how first ten rows look like.
```{r group}
grouped_mydata<-aggregate(mydata$steps,by=list(date=mydata$date),FUN=sum,na.rm=TRUE)
head(grouped_mydata,10)
```
##Make a histogram of the total number of steps taken each day

Here is a histogram:
```{r hist}
hist(grouped_mydata$x)
```

##Calculate and report the mean and median of the total number of steps taken per day

```{r mean,results='hide'}
steps_mean<-(mean(grouped_mydata$x,na.rm = TRUE))
steps_mean_round<- format (steps_mean,digits=5)
```

```{r median, results='hide'}
steps_median<-(median(grouped_mydata$x, na.rm = TRUE))
```
Mean total number of steps taken per day is `r steps_mean_round`

Median total number of steps taken per day is `r steps_median`


##What is the average daily activity pattern?

Let's re-group our original data file and see how top ten row look like:

```{r re-group}
grouped_mydata_time<-aggregate(mydata$steps,by=list(interval=mydata$interval),FUN= mean, na.rm=TRUE)
head(grouped_mydata_time,10)
```

##Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r plot}
plot(grouped_mydata_time$interval,grouped_mydata_time$x)
```

##Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max steps,results='hide'}
grouped_mydata_time[which.max(grouped_mydata_time$x),]
GMT<-grouped_mydata_time[which.max(grouped_mydata_time$x),]
GMT_I<-GMT[1,1]
GMT_x<-round(GMT[1,2],digits = 2)

```
Interval `r GMT_I` contains max number of steps - `r GMT_x`.


##Inputing missing values

##Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r count NA, results='hide'}
count_NA<-sum(is.na(mydata$steps))
```
Total number of rows with NAs is `r count_NA`.

##Filling in all of the missing values in the dataset. 

I'm agoing to use the mean for that 5-minute interval.

Let's make sure data in the main data file is sorted first by date and then by interval.
```{r check order1}
neworder<-mydata[order(mydata[,2],mydata[,3]),]
head(neworder,10)
```
Now, let's make sure that file contains totals by intervals is sorted by intervals.

```{r check order2}
neworder2<-grouped_mydata_time[order(grouped_mydata_time[,1],grouped_mydata_time[,2]),]
head(neworder2)
```

Let's add a new column with daily average for each interval

```{r  adding average}
neworder2$average_steps<-round(neworder2$x/61,digits = 2)
head(neworder2)
```
Let's merger these two correctly ordered files:
```{r merger}
merger_data<-merge(neworder,neworder2)
head(merger_data)
```
Let's just check dimentions of the new merged file:
```{r, check dim}
dim(neworder)
dim(neworder2)
dim(merger_data)
```
Let's re-order these data my date and interval
```{r reorder merger_data}
neworder3<-merger_data[order(merger_data[,3],merger_data[,1]),]
head(neworder3)
```
Finally, let's replace all NAs with data from Average_steps column. 

So, the logic is if the data is not available for a particular date and particular interval, we will use two months average for this interval as approximation.

```{r replace NAs}
neworder3$steps[is.na(neworder3$steps)]<-neworder3$average_steps
head(neworder3)
```
##Creating new data set

Now, I will create a new dataset that is equal to the original dataset but with the missing data filled in. Let's take a look at first ten records:

```{r new data set}
new_data_set<-data.frame(neworder3$steps,neworder3$date,neworder3$interval)
names(new_data_set)<-c("steps","date","interval")
head(new_data_set,10)
```

##Calculating new numbers and making new histogram.

Here is a histogram of the total number of steps taken each day.
```{r group by date - no NA}
grouped_new_data_set<-aggregate(new_data_set$steps,by=list(date=new_data_set$date),FUN=sum)
head(grouped_new_data_set,10)
```

```{r hist - no NA}
hist(grouped_new_data_set$x)
```



Here I re-calculate and report the mean and median total number of steps taken per day. 

```{r mean  - no NAs}
steps_mean_nona<-(mean(grouped_new_data_set$x))
steps_mean_nona_round<- format (steps_mean_nona,digits=5)
```

```{r median - no NAs}
steps_median_nona<-(median(grouped_new_data_set$x))
steps_median_nona_round<- format (steps_median_nona,digits=5)
```
Mean total number of steps taken per day is `r steps_mean_nona_round` vs.`r steps_mean_round` before NAs were removed.

Median total number of steps taken per day is `r steps_median_nona_round` vs. `r steps_median` before NAs were removed.





##Are there differences in activity patterns between weekdays and weekends?
I will creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r add weekdays}
new_data_set$weekdays<-weekdays(as.Date(new_data_set$date))
head(new_data_set,10)
```


Instead of displaying days of the week, I want to see "Yes" and "No" in the column

```{r adding yes and no}
weekdays = c ("Monday","Tuesday", "Wednesday","Thursday","Friday", "Saturday", "Sunday")
DN = c(1,2,3,4,5,6,7)
Weeks=data.frame(weekdays,DN)
head(Weeks)

merger_day<-merge(new_data_set,Weeks)
head(merger_day)

merger_day<-merger_day[order(merger_day$date,merger_day$interval),]
head(merger_day)
```

```{r assign yes or no}
merger_day$Weekend<-ifelse(merger_day$DN > 5,"Weekend","Weekday")
head(merger_day)
```
```{r}
merger_day<-merger_day[c(2,3,4,6)]
head(merger_day)
```

##Making a new plot
First, let's break the 'weekend' file in two
Weekend portion:
```{r weekend portion}
grouped_by_w<-subset(merger_day,merger_day$Weekend == "Weekend")
head (grouped_by_w,10)
```
Weekday portion:
```{r non-weekend portion}
grouped_by_nw<-subset(merger_day,merger_day$Weekend == "Weekday")
head (grouped_by_nw,10)
```


##Ploting graphs
###Here is 'weekend' portion

```{r plot weekend graphs,echo = FALSE}
plot(grouped_by_w$interval,grouped_by_w$steps,type="l")
```


###Here is 'weekday' portion
```{r plot weekend graphs2,echo = FALSE}
plot(grouped_by_nw$interval,grouped_by_nw$steps,type="l")
```

###The End